A family of accessibility measures derived from spatial interaction principles

### 2025-08-05 paper_plos/paper_plos.tex
made some additional, smaller, cuts. still 37 pages.

More specifically, the wide range of accessibility definitions--with
novel methods being more sophisticated but less intuitive
\citep{kapatsila_resolving_2023}--can further hinder practical uptake
\citep{vanweeAccessible2016}. Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004} classify accessibility
measures into four categories: infrastructure-, place-, person-, and
utility-based. Among place-based measures (this work's focus), variants
include gravity-based \citep{hansen1959}, cumulative opportunity
\citep{pirie_measuring_1979}, the 2 Step Floating Catchment Area (FCA)
approach \citep{luo2003}, and a variety of modifications to these
approaches e.g., Enhanced 2-Step FCA \citep{luoEnhanced2009}, 3-Stage
FCA \citep{wan2012three}, Modified 2-Step FCA
\citep{delamater2013spatial}, inverse 2-Step FCA
\citep{wang_2sfca_2021}, and n-steps FCA \citep{liang_novel_2024}. While
these methods are tailored to address specific research contexts,
overall this diversity does not demystify existing questions like those
raised in \citet{vanweeAccessible2016}: How should practitioners
interpret difference in scores between modes, and how should results be
communicated?

@@

We contend that accessibility research should reengage with spatial
interaction modeling, particularly the use of Wilson's system
constraints. While spatial interaction models embraced such constraints
to improve interpretability, most accessibility models have not. This
lack of adoption contributes to fuzziness in current analyses, limiting
interpretive clarity to simple proportional comparisons (e.g.,
``higher-than,'' ``lower-than'')
\citep{millerAccessibilityMeasurementApplication2018}. Without such
constraints, accessibility scores lack clear units and comparability
across cities or modes. In contrast, constrained measures yield values
that can be tied to tangible values without any post-hoc treatment,
making them more interpretable, communicable, and actionable in
planning.

This paper lays the historical and theoretical groundwork for a
companion empirical study demonstrating how these constrained forms
restore interpretability through concrete measurement units (e.g.,
number of accessible opportunities), as opposed to the abstract weighted
outputs typical of conventional gravity models.

Accessibility and spatial interaction modelling literature share common
headwaters, and the latter has given careful attention to measurement
units and their interpretability. Following this introduction, this
paper revisits that lineage--from Ravenstein's Newtonian gravitational
expression of spatial interaction \citep{ravensteinLawsMigration1889},
through Hansen \citep{hansen1959}, to Wilson \citep{wilson1971}--to
propose a unified family of constrained accessibility measures. We
illustrate each case of the family with a numerical toy example and
conclude by discussing how these constrained measures can inform
planning decisions and improve clarity in accessibility analysis.

@@

The patterns of people's movement in space have been a subject of
scientific inquiry for at least a century and a half. From as far back
as Henry C. Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858}, a concern with the scientific
study of human spatial interaction can be observed. It was in this work
where Carey stated that ``man {[}is{]} the molecule of society {[}and
their interaction is subject to{]} the direct ratio of the mass and the
inverse one of distance'' \citep[pp.~37-38]{mckeanManual1883}. This
statement shows how investigations into human spatial interaction have
often been explicitly coloured by the features of Newton's Law of
Universal Gravitation, first posited in 1687's \emph{Principia
Mathematica} and expressed as in Eq~\ref{eq-phys-grav-prop}.

@@

This famous equation expresses that the attractive force \(F\) between
two bodies is directly proportional to the product of their masses and
inversely proportional to the square of the distance between them. As
mass increases, so does force; as distance increases, force decreases.
However, Eq~\ref{eq-phys-grav-prop} only demonstrates a proportional
relationship. To quantify the magnitude (not the proportional magnitude)
of the force, it must be \emph{constrained} with an empirical constant.
This constant \(G\) converts Eq~\ref{eq-phys-grav-prop} from an
expression of proportionality to the following expression of equality:

@@

Where \(G\) is the gravitational constant--an empirically calibrated
value that ensures the model reflects observed forces. Newton's initial
estimate of \(G\) was based on a speculation that the mean density of
earth was between five or six times that of water, an assumption that
received support after Hutton's experiments of 1778
\citep[p.~783]{hutton_xxxiii_1778}. Still, it took over a century from
the publication of \emph{Principia} to refine the estimate of the
proportionality constant to within 1\% accuracy, with Cavendish's 1798
experiment \citep{cavendish_xxi_1798}.

This gravitational framing laid the conceptual groundwork for later
empirical studies of human spatial interaction, such as those by
Ravenstein, Stewart, and others, who began translating these principles
into observed patterns of human movement.

@@

These early formulations (Eq~\ref{eq-reilly}, Eq~\ref{eq-zipf}) clearly
reflect the influence of Newtonian gravity on human spatial interaction
theory, revealing a shared mathematical structure across migration,
trade, and communication models. However, none included a
proportionality constant--highlighting a consistent omission of the
empirical calibration necessary to convert these proportional
relationships into measurable and comparable quantities

After Carey and folowing Reilly and Zipf, it is only in
\citet{stewartDemographicGravitationEvidence1948} that we find the most
explicit connection yet to Newton's Gravitational law and the use of a
proportionality constant. Besides awareness of preceding research (i.e.,
Stewart cites both Reilly and Zipf as predecessors in the analysis of
human spatial interaction), Stewart appears to have been the first
author to express his theorized relationships for human spatial
interaction with a proportionality constant \(G\), enabling his
formulation to be interpreted as a measurable `demographic' force:

@@

\noindent Where \(M_j\) is the demographic mass at location \(j\) and
\(d_{ij}\) is the distance between \(i\) and \(j\). A version of this
discrete form is what was used in \citet{hansen1959}, going on to become
a foundation of modern accessibility definitions, as will be discussed.

Although Stewart's concept of ``social physics'' eventually fell out of
favour--due to inconsistent mathematical notation (e.g., \(G\) is used
as both a proportionality constant p.~34 and then later as
\emph{demographic energy} on p.~53.) and its foundation in racist and
unscientific assumptions (e.g., view of humans as particles following
physical laws and assumptions of the molecular weight \(\mu\) of the
average American being one, but ``presumably\ldots much less than
one\ldots.for an Australian aborigine'' {[}p.~35{]})--his introduction
of a proportionality constant \(G\) in modeling demographic force marks
an important conceptual step: recognizing that moving from
proportionality to equality requires empirical calibration. In other
words, the addition of \(G\) shifts results from being abstract
indicators of potential (i.e., people\textsuperscript{2-distance}-2) to
having interpretable units grounded in consistent, albeit still
abstract, quantities (i.e., units of demographic force).

As will be discussed, when Hansen later adopted Stewart's formula
(Eq~\ref{eq-stewart-population-potential-sum}) for accessibility, he
omitted any mention of \(G\), effectively setting it to 1 arbitrarily as
Stewart had done. This omission has persisted in accessibility research,
leaving a conceptual gap in how such measures are interpreted and
compared.

@@

From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
arrive to 1959 and Walter G. Hansen, whose work proved to be
exceptionally influential in the accessibility literature
\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
as ``the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart
\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
being a student of city and regional planning at the Massachusetts
Institute of Technology, Hansen was also an engineer with the Bureau of
Roads, and preoccupied with the power of transportation to shape land
uses in a very practical sense. \citet{hansen1959} drew directly from
Stewart's population potential (see
Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
broader (and often problematic) framework of ``social physics.''.

Hansen recast Stewart's population potential to reflect accessibility, a
model of human behaviour useful to capture regularities in mobility
patterns. Hansen replaced \(M_j\) in
Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
to derive an \emph{opportunity potential}, or more accurately, a
\emph{potential of opportunities for interaction} as
\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
rewriting to accounts for a variety of impedance functions beyond the
inverse power \(d^{-\beta}\) as follows:

@@

However, the omission of a proportionality constants and working with a
proportional relationship generates fundamental issues in comparability
between and, arguably within, studies. As the absence of \(G\) means
accessibility estimates have no fixed unit--this renders them sensitive
to the choice of impedance functions. For instance, if travel cost
\(d_{ij}\) is measured in meters, then when the travel impedance
function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the resulting \(S_i\)
has units of opportunities per \(\text{metres}^{\beta}\); but when
\(f(d_{ij})\) is set to equal \(e^{-\beta d_{ij}}\), the units become
opportunities per \(e^{\beta \text{metres}}\). Such variation impairs
comparability across analysis and obscures the meaning of accessibility
scores renders their results difficult to understand and communicate
without post-hoc treatment.

Therefore in practice, Hansen-type measures (of proportionality) are
better understood as \emph{ordinal indicators}--they rank accessibility
but lack cardinal meaning or consistent units
\citep{millerAccessibilityMeasurementApplication2018}. The continued
absence of \(G\) leaves a conceptual and practical gap in accessibility
analysis: a missing link between theoretical form and empirical
measurement.

@@

While accessibility research evolved in North America with
\citet{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. In his groundbreaking study
\citep{wilson1971}, advanced a general framework for spatial interaction
modeling--focused not on potential (as accessibility), but on modeling
observed flows of interaction between places. \citet{wilson1971}
formalized the general spatial interaction model through the following
equation:

@@

The model in Eq~\ref{eq-phys-gravity-model} posits a quantity \(T_{ij}\)
that represents a value in a matrix of flows of size \(n \times m\),
that is, between \(i = 1,s, n\) origins and \(j = 1,s, m\) destinations.
The quantities \(W_i^{(1)}\) and \(W_j^{(2)}\) are proxies for the
masses at \(i=1,s,n\) origins and \(j=1,s,m\) destinations. The
super-indices \((1)\) and \((2)\) indicate that these masses can be any
number of different things associated with the zones, i.e.,
\(W_i^{(1)}\) could be population at a zone as an origin, and
\(W_j^{(2)}\) hectares of park space at a zone as a destination.
\(f(c_{ij})\) is some function of travel cost \(c_{ij}\) which reflects
travel impedance. Finally, \(k\) acts as a proportionality constant,
shifting the equation from a proportional to an equal relationship by
incorporating known system totals. In this specific sense, we infer that
\(k\) serves a role similar to the gravitational constant in Newton's
law--it calibrates the model so that outputs match real-world
quantities.

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) is represented as \(O_i\) and the following constraint can
be used:

@@

It is also possible to have information about both \(O_i\) and \(D_j\),
in which case both constraints could be imposed on the model at once.

Depending on which system constraints are applied---fully, partially, or
not at all---a family of spatial interaction models can be derived from
Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
replaced with different \emph{balancing factors}. This change in name is
more useful as these factors not only preserve proportionality but also
ensure that the predicted flows \(T_{ij}\) align with the known
constraints considered in the system. In other words, the different
balancing factors adjust the model so that flows become statistical
averages consistent with observed origin and/or destination data. In the
framework introduced and inferred from \citet{wilson1971}, three types
of balancing factors are specified: (1) an unconstrained model that only
matches the total volume of interaction \(K\), (2) a singly constrained
model (either by origins \(A_i\) or destinations \(B_j\)), and (3) a
doubly constrained model that satisfies both.

In the unconstrained model, constraints in
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
this means that the total number of flows predicted by the model must be
equal to sum of all flows from origins \(i\) to destinations \(j\). The
balancing factor \(K\) takes the place of \(k\) and is equal to the
following \citep[as specified in][ and
\citet{fotheringham_spatial_1984}]{cliff_evaluating_1974}:

@@

In the singly-constrained model, only constraint
Eq~\ref{eq-constraint1-gravitymodel} or constraint
Eq~\ref{eq-constraint2-gravitymodel} hold. When only
Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
to the following production-constrained singly-constrained model:

\begin{equation}\phantomsection\label{eq-production-constrained-gravitymodel}{
T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
}\end{equation}

Notice how, in this model, the proxy for the mass at the origin
\(W_i^{(1)}\) is replaced with \(O_i\), representing what we know about
the system, the spatial interaction outbound flow, i.e., outbound flows
produced at \(i\). Also, there is no longer a single system-wide
balancing factor \(K\), but rather a set of balancing factors specific
to origins \(A_i\).

For this model, the balancing factors ensure that constraint
Eq~\ref{eq-constraint1-gravitymodel} is satisfied, meaning that the sum
of predicted flows from one origin going to all destinations must equal
the known mass at that origin \(O_i\) e.g., the total number of outbound
flows. Satisfying this constraint also implicitly fulfills the total
constraint (Eq~\ref{eq-constraint0-gravitymodel}), since the sum of
\(O_i\) values across all origins equals the total number of flows.

The production-constrained singly-constrained model is useful when flows
ends are unknown but the number of flows originating from each location
is known. The balancing factors for the production-constrained model are
solved for each origin \(A_i\), and according to Wilson are:

@@

The attraction-constrained model is similar to the
production-constrained model as it is also singly constrained but from
the perspective of the mass at the destination. From the
attraction-constrained model, the proxy for the mass at the destination
\(W_j^{(2)}\) is now replaced with \(D_j\), representing the spatial
interaction inbound flow e.g., flows attracted to the destination, and
takes the following form:

\begin{equation}\phantomsection\label{eq-attraction-constrained-gravitymodel}{
T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
}\end{equation}

In this model (Eq~\ref{eq-attraction-constrained-gravitymodel}), the
balancing factors ensure that Eq~\ref{eq-constraint2-gravitymodel} is
satisfied (hence the total constraint
Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning that the sum
of predicted flows going to one destination from all origins must equal
the known mass of that destination \(D_j\) i.e., the total number of
inbound flows to \(j\). This should hold for all destinations. As
before, destination-specific balancing factors \(B_j\) were derived by
Wilson as:

@@

Lastly, the doubly-constrained model is the production-attraction
constrained model in \citet{wilson1971}. In this case, both constraints
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and takes
the following form:

\begin{equation}\phantomsection\label{eq-doubly-constrained-gravitymodel}{
T_{ij} = A_i B_j O_i D_j f(c_{ij})
}\end{equation}

In this model, both proxies for the masses are replaced with the known
masses, that is, the flows produced by the origin and the flows
attracted by the destination. There are now two sets of mutually
dependent proportionality constants shown in
Eq~\ref{eq-doubly-constrained-balancing-factors}. Derivation of these
models is demonstrated in detail elsewhere
\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.

@@

Wilson's work is notable for many reasons, one of which is his departure
from the metaphor of Newtonian gravitational physics. Instead, he
grounded his models in statistical mechanics, particularly the principle
of entropy maximisation. For accessibility research, Wilson's approach
is especially significant because it can address a problem that has
quietly plagued this field (as this work has argued): the transition
from proportionality to equality.

Rather than relying on a universal constant like \(G\), Wilson's models
calibrate interaction flows through known empirical constraints (e.g.,
total system flows, or origin-side and destination-side flow totals).
This results in interpretable, balanced (given the constraints), and
unit-consistent \(ij\) flows. In short, when aspects of the systems are
inputs into the balacing factors, \(ij\) flows can be robustly predicted
in tangible units of `flow', units that align with the system being
modelled. In this way, spatial interaction modeling tradition can be
seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today)
yield outputs that \emph{indicate} a proportional relationship and not
one of equality nor yielding results without interpretable units.

@@

In this way, these balancing factors can be used to recover what
Stewart's \(G\) was meant to accomplish, but through a different
conceptual path. But before demonstrating a family of accessibility
measures that are sensitive to constraints, in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach focused, until this work.

@@

This divergence is especially striking given the context in which both
frameworks emerged. As noted in
\citet{battyChronicleScientificPlanning1994}, large scale spatial
interaction models (like Wilson's) responded to important developments
at the time, a need ``to meet the dictates and needs of public policy
for strategic land use and transportation planning''. And these needs
were far from trivial: in the U.S., for instance, the Federal-Aid
Highway Act of 1956 set in motion the construction of the Interstate
Highway System with an eventual budget exceeding \$600 billion in
today's dollars
\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
this context, spatial interaction models were incorporated into
institutional practices focused on ``predict and provide'' travel demand
forecasting
\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
Accessibility analysis, by contrast, remained more conceptually diffuse,
focused on indicators of ``potential'' spatial interaction with
opportunities rather than flows that could tangibly guide infrastructure
decisions (e.g., roadway capacity expansion, new construction). Whereas
spatial interaction modelling became a key element of transportation
planning practice, accessibility remained a somewhat more academic
pursuit, and the two streams of literature only rarely connected.

>>>>

More specifically, the wide range of accessibility definitions--with
novel methods being more sophisticated but less intuitive
\citep{kapatsila_resolving_2023}--can further hinder practical uptake
\citep{vanweeAccessible2016}. Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004} classify accessibility
measures into four categories: infrastructure-, place-, person-, and
utility-based. Among place-based measures (this work's focus), variants
include gravity-based \citep{hansen1959}, cumulative opportunity
\citep{pirie_measuring_1979}, the 2 Step Floating Catchment Area (FCA)
approach \citep{luo2003}, and a variety of modifications to these
approaches e.g., Enhanced 2-Step FCA \citep{luoEnhanced2009}, 3-Stage
FCA \citep{wan2012three}, Modified 2-Step FCA
\citep{delamater2013spatial}, inverse 2-Step FCA
\citep{wang_2sfca_2021}, and n-steps FCA \citep{liang_novel_2024}. While
these methods are tailored to address specific research contexts,
overall this diversity does not demystify existing questions like those
raised in \citep{vanweeAccessible2016}: How should practitioners
interpret difference in scores between modes, and how should results be
communicated?

@@

We contend that accessibility research should re-engage with spatial
interaction modeling, particularly the use of Wilson's system
constraints. While spatial interaction models embraced such constraints
to improve interpretability, most accessibility models have not. This
lack of adoption contributes to fuzziness in current analyses, limiting
interpretive clarity to simple proportional comparisons (e.g.,
``higher-than'', ``lower-than'')
\citep{millerAccessibilityMeasurementApplication2018}. Without such
constraints, accessibility scores lack clear units and comparability
across cities or modes. In contrast, constrained measures yield values
that can be tied to tangible values without any post-hoc treatment,
theoretically making them more interpretable, communicable, and
actionable in planning.

Accessibility and spatial interaction modelling literature share common
headwaters, and the latter has given careful attention to measurement
units and their interpretability. Following this introduction, this
paper revisits that lineage--from Ravenstein's Newtonian gravitational
expression of spatial interaction \citep{ravensteinLawsMigration1889},
through Hansen \citep{hansen1959} to Wilson \citep{wilson1971}--to
propose a unified family of constrained accessibility measures. We
illustrate each case of the family with a numerical toy example and
conclude by discussing how these constrained measures can inform
planning decisions and improve clarity in accessibility analysis.

@@

The patterns of people's movement in space have been a subject of
scientific inquiry for at least a century and a half, from as far back
as Henry C. Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858}. It was in this work where
Carey stated that ``man {[}is{]} the molecule of society {[}and their
interaction is subject to{]} the direct ratio of the mass and the
inverse one of distance'' \citep[pp.~37-38]{mckeanManual1883}. This
statement shows how investigations into human spatial interaction have
often been explicitly coloured by the features of Newton's Law of
Universal Gravitation, first posited in 1687's \emph{Principia
Mathematica} and expressed as in Eq~\ref{eq-phys-grav-prop}.

@@

This famous equation expresses that the attractive force \(F\) between
two bodies is directly proportional to the product of their masses and
inversely proportional to the square of the distance between them. As
mass increases, so does force; as distance increases, force decreases.
However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
\emph{proportional} relationship. To quantify the magnitude (not the
\emph{proportional} magnitude) of the force, it must be
\emph{constrained} with an empirical constant. This constant \(G\)
converts Eq~\ref{eq-phys-grav-prop} from an expression of
proportionality to the following expression of equality:

@@

Where \(G\) is the gravitational constant--an empirically calibrated
value that ensures the model reflects observed forces. Newton's initial
estimate of \(G\) was based on a speculation that the mean density of
earth was between five or six times that of water, an assumption that
received support after Hutton's experiments of 1778
\citep[p.~783]{hutton_xxxiii_1778} and Cavendish's 1798 experiment
\citep{cavendish_xxi_1798} to estimate of the proportionality constant
to within 1\% accuracy. In sum: it took over a century from the
publication of \emph{Principia} to refine the estimate of the
proportionality constant.

The Newtonian gravitational relationship laid the conceptual groundwork
for later empirical studies of human spatial interaction, such as those
by Ravenstein, Stewart, and others, who began translating these
principles into observed patterns of human movement. However, the
majority of these attempts described a proportional relationship, or one
arbitrarily set to equality.

@@

These early formulations (Eq~\ref{eq-reilly}, Eq~\ref{eq-zipf}) clearly
reflect the influence of Newtonian gravity on human spatial interaction
theory, revealing a shared mathematical structure across migration,
trade, and communication models. However, none included a
proportionality constant--highlighting a consistent omission of the
empirical calibration necessary to convert these proportional
relationships into measurable and comparable quantities.

After Carey and folowing Reilly and Zipf, it is only in Stewart
\citep{stewartDemographicGravitationEvidence1948} that we find the most
explicit connection yet to Newton's Gravitational law and the use of a
proportionality constant. Besides awareness of preceding research (i.e.,
Stewart cites both Reilly and Zipf as predecessors in the analysis of
human spatial interaction), Stewart appears to have been the first
author to express his theorized relationships for human spatial
interaction with a proportionality constant \(G\), enabling his
formulation to be interpreted as a measurable `demographic' force:

@@

\noindent Where \(M_j\) is the demographic mass at location \(j\) and
\(d_{ij}\) is the distance between \(i\) and \(j\). A version of this
discrete form is what was used in Hansen \citep{hansen1959}, going on to
become a foundation of modern accessibility definitions, as will be
discussed.

Although Stewart's concept of ``social physics'' eventually fell out of
favour--due to inconsistent mathematical notation (e.g., \(G\) is used
as both a proportionality constant p.~34 and then later as
\emph{demographic energy} on p.~53.) and its racist and unscientific
assumptions (e.g., view of humans as particles following physical laws
and assumptions of the molecular weight \(\mu\) of the average American
being one, but ``presumably\ldots much less than one\ldots.for an
Australian aborigine'' {[}p.~35{]})--his introduction of a
proportionality constant \(G\) in modeling demographic force marks an
important conceptual step: recognizing that moving from proportionality
to equality requires empirical calibration. In other words, the addition
of \(G\) shifts results from being abstract indicators of potential
(i.e., \(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having
interpretable units grounded in consistent, albeit still abstract,
quantities (i.e., units of demographic force).

As will be discussed, when Hansen \citep{hansen1959} later adopted
Stewart's formula (Eq~\ref{eq-stewart-population-potential-sum}) for
accessibility, he omitted any mention of \(G\), effectively setting it
to 1 arbitrarily as Stewart had done. This omission has persisted in
accessibility research, leaving a conceptual gap in how such measures
are interpreted and compared.

@@

From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
arrive to 1959 and Walter G. Hansen, whose work proved to be
exceptionally influential in the accessibility literature
\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
as ``the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart
\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
being a student of city and regional planning at the Massachusetts
Institute of Technology, Hansen was also an engineer with the Bureau of
Roads, and preoccupied with the power of transportation to shape land
uses in a very practical sense. Hansen \citep{hansen1959} drew directly
from Stewart's population potential formula (see
Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
broader (and often problematic) framework of ``social physics''.

Hansen recast Stewart's population potential to reflect accessibility, a
model of human behaviour useful to capture regularities in mobility
patterns. Hansen replaced \(M_j\) in
Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
to derive an \emph{opportunity potential}, or more specifically, a
\emph{potential of opportunities for interaction} as
\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
rewriting to accounts for a variety of impedance functions beyond the
inverse power \(d^{-\beta}\) as follows:

@@

Furthermore, working with a proportional relationship generates
fundamental issues in comparability between and, arguably within,
studies. Namely, accessibility estimates have no fixed unit, rendering
them sensitive to the choice of impedance functions. For instance, if
travel cost \(d_{ij}\) is measured in meters, then when the travel
impedance function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the
resulting \(S_i\) has units of opportunities per
\(\text{metres}^{\beta}\). However, when \(f(d_{ij})\) is set to equal
\(e^{-\beta d_{ij}}\), the units become opportunities per
\(e^{\beta \text{metres}}\). Such variation impairs comparability across
analysis and obscures the meaning of accessibility scores, making them
difficult to understand and communicate without post-hoc treatment.

Therefore in practice, Hansen-type measures are ones of proportionality
and are better understood as \emph{ordinal indicators}; they rank
accessibility but lack cardinal meaning or consistent units
\citep{millerAccessibilityMeasurementApplication2018}. The continued
absence of a sort of proportionality constant \(G\) leaves a conceptual
and practical gap in accessibility analysis: a missing link between
theoretical form and empirical measurement.

@@

While accessibility research evolved in North America with Hansen
\citep{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. Wilson's ground breaking paper
\citep{wilson1971}, advanced a general framework for spatial interaction
modeling--focused on modeling flows of interaction between places. This
work was not focused on the `potential' concept as accessibility has
been. Wilson \citep{wilson1971} formalized the general spatial
interaction model through the following equation:

@@

The model in Eq~\ref{eq-phys-gravity-model} posits a quantity \(T_{ij}\)
that represents a value in a matrix of flows of size \(n \times m\),
that is, between \(i = 1,s, n\) origins and \(j = 1,s, m\) destinations.
The quantities \(W_i^{(1)}\) and \(W_j^{(2)}\) are proxies for the
masses at \(i=1,s,n\) origins and \(j=1,s,m\) destinations. The
super-indices \((1)\) and \((2)\) indicate that these masses can be any
number of different things associated with the zones, i.e.,
\(W_i^{(1)}\) could be population at a zone as an origin, and
\(W_j^{(2)}\) hectares of park space at a zone as a destination.
\(f(c_{ij})\) is some function of travel cost \(c_{ij}\) which reflects
travel impedance.

Of important note for this paper, \(k\) in
Eq~\ref{eq-phys-gravity-model} acts as a proportionality constant,
shifting the equation from a proportional to an equal relationship by
incorporating known system totals. In this sense, we infer that \(k\)
serves a role similar to the gravitational constant in Newton's law--it
calibrates the model so that outputs match real-world quantities.
However, these real-world quantities was not a set empirical constant
(like Newton's \(G\)) but instead, sensitive to the system and known
information about the system.

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model} is represented as
\(O_i\) and the following constraint can be used:

@@

It is also possible to have information about both \(O_i\) and \(D_j\),
in which case both constraints can be imposed on the model at once.

Depending on which of the three system constraints are applied a family
of spatial interaction models can be derived from
Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
replaced with different \emph{balancing factors}. This change in name is
more useful as these factors not only preserve proportionality but also
ensure that the predicted flows \(T_{ij}\) align with the known
constraints considered in the system. In other words, the different
balancing factors adjust the model so that flows become statistical
averages consistent with observed origin and/or destination data. In the
framework introduced and inferred from Wilson \citep{wilson1971}, three
types of balancing factors are specified: (1) an unconstrained model
that only matches the total volume of interaction \(K\), (2) a singly
constrained model (either by origins \(A_i\) or destinations \(B_j\)),
and (3) a doubly constrained model that satisfies both.

In the unconstrained model, constraints in
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
this means that the total number of flows predicted by the model must be
equal to sum of all flows from origins \(i\) to destinations \(j\). The
balancing factor \(K\) takes the place of \(k\) and is equal to the
following (as specified in \citep{cliff_evaluating_1974} and
\citep{fotheringham_spatial_1984}):

@@

In the singly-constrained model, only constraint
Eq~\ref{eq-constraint1-gravitymodel} or constraint
Eq~\ref{eq-constraint2-gravitymodel} hold. When only
Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
to the production-constrained singly-constrained version of
Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
origin \(W_i^{(1)}\) is replaced with \(O_i\). Also, \(k\) is replaced
with a set of balancing factors specific to origins \(A_i\) which
ensures that constraint Eq~\ref{eq-constraint1-gravitymodel} is
satisfied (i.e., the sum of predicted flows from one origin going to all
destinations must equal the known mass at that origin \(O_i\)).
Satisfying this constraint also implicitly fulfills the total constraint
(Eq~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\) values
across all origins equals the total number of flows. \(A_i\) takes the
following form:

@@

The singly-constrained attraction-constrained model is similar to the
production-constrained version but from the perspective of the mass at
the destination. For the attraction-constrained model, the proxy for the
mass at the destination \(W_j^{(2)}\) is replaced with \(D_j\) in
Eq~\ref{eq-phys-gravity-model}, representing the spatial interaction
inbound flow. Also, \(k\) is replaced with a set of destination-specific
balancing factors \(B_j\) that ensure that
Eq~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
constraint Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning
that the sum of predicted flows going to one destination from all
origins must equal the known mass of that destination \(D_j\). As
before, destination-specific balancing factors \(B_j\) were derived by
Wilson as:

@@

Lastly, the doubly-constrained model is the production-attraction
constrained model in Wilson \citep{wilson1971}. In this case, both
constraints Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and from
Eq~\ref{eq-phys-gravity-model}, \(k\) becomes both \(A_i\) and \(B_j\)
shown in Eq~\ref{eq-doubly-constrained-balancing-factors}, and
\(W_i^{(1)}\) and \(W_j^{(2)}\) is replaced with \(O_i\) and \(D_j\).
Derivation of these models is demonstrated in detail elsewhere
\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.

@@

Wilson's work is notable for many reasons, one of which is his departure
from the metaphor of Newtonian gravitational physics. Instead, he
grounded his models in statistical mechanics, particularly the principle
of entropy maximisation. For accessibility research, Wilson's approach
is especially significant because it can address a problem that has
quietly afflicted this field (as this paper has argued): the transition
from proportionality to equality.

Rather than relying on a universal constant like \(G\), Wilson's models
calibrate interaction flows through known empirical constraints (e.g.,
total system flows, or origin-side and destination-side flow totals).
This results in interpretable, balanced (given the constraints), and
unit-consistent \(ij\) flows. In short, when aspects of the systems are
inputs into the balancing factors, \(ij\) flows can be robustly
predicted in tangible units of `flow', units that align with the system
being modelled. In this way, spatial interaction modeling tradition can
be seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today)
yield outputs that \emph{indicate} a proportional relationship and not
one of equality nor yielding results without interpretable units.

@@

In this way, these balancing factors can be used to recover what
Stewart's \(G\) was meant to accomplish, but through a different
conceptual path. But before demonstrating a family of accessibility
measures that are sensitive to constraints, in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach focused, until this paper.

@@

This divergence is especially striking given the context in which both
frameworks emerged. As noted in
\citep{battyChronicleScientificPlanning1994}, large scale spatial
interaction models (like Wilson's) responded to important developments
at the time, a need ``to meet the dictates and needs of public policy
for strategic land use and transportation planning''. And these needs
were far from trivial: in the U.S., for instance, the Federal-Aid
Highway Act of 1956 set in motion the construction of the Interstate
Highway System with an eventual budget exceeding \$600 billion in
today's dollars
\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
this context, spatial interaction models were incorporated into
institutional practices focused on ``predict and provide'' travel demand
forecasting
\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
Accessibility analysis, by contrast, remained more conceptually diffuse,
focused on indicators of ``potential'' spatial interaction with
opportunities rather than flows that could tangibly guide infrastructure
decisions (e.g., roadway capacity expansion, new construction). Whereas
spatial interaction modelling became a key element of transportation
planning practice, accessibility remained a somewhat more academic
pursuit, and the two streams of literature only rarely connected.
